---
title: "mzmlr Introduction"
author: "mzmlr Developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mzmlr Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `mzmlr` package provides functionality to validate and read mzML files, the standard format for mass spectrometry data defined by HUPO-PSI. Unlike the `mzR` package which relies on external C++ libraries, `mzmlr` uses native R XML parsing capabilities via the `xml2` package.

## Installation

```{r installation, eval = FALSE}
# Install from CRAN
install.packages("mzmlr")

# Or install development version
# devtools::install_github("yourusername/mzmlr")
```

## Basic Usage

### Reading an mzML File

```{r read-example, eval = FALSE}
library(mzmlr)

# Read an mzML file with validation
mzml <- read_mzml("path/to/file.mzML")

# The returned object is of class MzMlFile
class(mzml)
# [1] "MzMlFile"
```

### File Validation

Before reading, you can validate an mzML file against the official XSD schema:

```{r validate-example, eval = FALSE}
result <- validate_mzml("path/to/file.mzML")

if (result$valid) {
  cat("Valid mzML version", result$version, "\n")
} else {
  cat("Validation errors:", result$message, "\n")
}
```

## Extracting Metadata

### File Information

```{r info-example, eval = FALSE}
info <- get_file_info(mzml)

# View summary
print(info)

# Access specific fields
info$spectrum_count
info$ms_levels
info$start_time
```

### Instrument Configuration

```{r instrument-example, eval = FALSE}
instr <- get_instrument_info(mzml)

# View instrument details
print(instr)

# Get components
instr$components
```

## Reading Spectrum Data

### Single Spectrum

```{r spectrum-example, eval = FALSE}
# Get the first spectrum
spec <- get_spectrum(mzml, index = 1)

# Plot the spectrum
plot(spec$mz, spec$intensity, type = "l",
     xlab = "m/z", ylab = "Intensity",
     main = paste("MS Level", spec$ms_level))
```

### Multiple Spectra

```{r spectra-example, eval = FALSE}
# Get first 10 spectra
spectra <- get_spectra(mzml, indices = 1:10)

# Get all MS1 spectra
ms1_spectra <- get_spectra(mzml, ms_level = 1)

# Process each spectrum
for (i in seq_along(ms1_spectra)) {
  spec <- ms1_spectra[[i]]
  cat(sprintf("Spectrum %d: %d peaks, TIC = %.2f\n",
              i, length(spec$mz), spec$total_ion_current))
}
```

## Reading Chromatogram Data

```{r chromatogram-example, eval = FALSE}
# Get all chromatograms
chromatograms <- get_chromatograms(mzml)

# Plot if available
if (length(chromatograms) > 0) {
  chrom <- chromatograms[[1]]
  plot(chrom$time, chrom$intensity, type = "l",
       xlab = "Time", ylab = "Intensity",
       main = chrom$chromatogram_type)
}
```

## Memory-Efficient Processing

For large files, process spectra one at a time to avoid memory issues:

```{r memory-efficient, eval = FALSE}
# Get total count without loading data
info <- get_file_info(mzml)
n_spectra <- info$spectrum_count

# Process in batches
batch_size <- 100
for (start in seq(1, n_spectra, by = batch_size)) {
  end <- min(start + batch_size - 1, n_spectra)
  batch <- get_spectra(mzml, indices = start:end)

  # Process batch
  # ...

  # Clear memory
  rm(batch)
  gc()
}
```

## Working with Binary Data

The package automatically handles:
- Base64 decoding
- zlib decompression
- 32-bit float and 64-bit double conversion

```{r binary-data, eval = FALSE}
# All binary data handling is automatic
spec <- get_spectrum(mzml, 1)

# Data is ready for analysis
summary(spec$mz)
summary(spec$intensity)
```

## Error Handling

The package provides informative error messages:

```{r error-handling, eval = FALSE}
tryCatch({
  mzml <- read_mzml("nonexistent.mzML")
}, error = function(e) {
  cat("Error:", conditionMessage(e), "\n")
})
```

## Advanced Features

### Custom Schema Validation

```{r custom-schema, eval = FALSE}
# Use a custom XSD schema for validation
result <- validate_mzml("file.mzML",
                        schema_path = "path/to/custom.xsd")
```

### Disabling Validation

For faster loading when validation is not needed:

```{r no-validation, eval = FALSE}
mzml <- read_mzml("file.mzML", validate = FALSE)
```

## See Also

- [HUPO-PSI mzML specification](https://www.psidev.info/mzml)
- [mzML GitHub repository](https://github.com/HUPO-PSI/mzML)
- [mzR package](https://cran.r-project.org/package=mzR)
