---
title: "mzmlr Introduction"
author: "mzmlr Developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mzmlr Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `mzmlr` package provides functionality to validate and read mzML files, the standard format for mass spectrometry data defined by HUPO-PSI. Unlike the `mzR` package which relies on external C++ libraries, `mzmlr` uses native R XML parsing capabilities via the `xml2` package.

## Installation

```{r installation, eval = FALSE}
# Install from CRAN
install.packages("mzmlr")

# Or install development version
# devtools::install_github("RECETOX/mzmlr")
```

## Reading mzML Files

### Local files

```{r read-local, eval = FALSE}
library(mzmlr)

# Read an mzML file with validation (default)
mzml <- read_mzml("path/to/file.mzML")

# The returned object is of class MzMlFile
class(mzml)
# [1] "MzMlFile"

# Disable validation for faster loading
mzml <- read_mzml("path/to/file.mzML", validate = FALSE)
```

### Reading from URLs

The package supports reading mzML files directly from web locations. The file is automatically downloaded to a temporary location and cleaned up when no longer needed:

```{r read-url, eval = FALSE}
# Read from a URL
mzml <- read_mzml("https://example.com/data/file.mzML")

# With custom timeout for large files
mzml <- read_mzml(
  "https://example.com/data/large_file.mzML",
  validate = TRUE
)

# The temp_file component contains the path to the downloaded file
# It will be automatically cleaned up when the R session ends
if (!is.null(mzml$temp_file)) {
  cat("Downloaded to:", mzml$temp_file, "\n")
}
```

### File Validation

Before reading, you can validate an mzML file against the official XSD schema:

```{r validate-example, eval = FALSE}
result <- validate_mzml("path/to/file.mzML")

if (result$valid) {
  cat("Valid mzML version", result$version, "\n")
} else {
  cat("Validation errors:", result$message, "\n")
}
```

## Extracting Metadata

### File Information

```{r info-example, eval = FALSE}
info <- get_file_info(mzml)

# View summary
print(info)

# Access specific fields
info$spectrum_count
info$ms_levels
info$start_time
```

### Instrument Configuration

```{r instrument-example, eval = FALSE}
instr <- get_instrument_info(mzml)

# View instrument details
print(instr)

# Get components
instr$components
```

## Reading Spectrum Data

### Single Spectrum

```{r spectrum-example, eval = FALSE}
# Get the first spectrum
spec <- get_spectrum(mzml, index = 1)

# Plot the spectrum
plot(spec$mz, spec$intensity,
  type = "l",
  xlab = "m/z", ylab = "Intensity",
  main = paste("MS Level", spec$ms_level)
)
```

### Multiple Spectra

```{r spectra-example, eval = FALSE}
# Get first 10 spectra
spectra <- get_spectra(mzml, indices = 1:10)

# Get all MS1 spectra
ms1_spectra <- get_spectra(mzml, ms_level = 1)

# Process each spectrum
for (i in seq_along(ms1_spectra)) {
  spec <- ms1_spectra[[i]]
  cat(sprintf(
    "Spectrum %d: %d peaks, TIC = %.2f\n",
    i, length(spec$mz), spec$total_ion_current
  ))
}
```

## Reading Chromatogram Data

```{r chromatogram-example, eval = FALSE}
# Get all chromatograms
chromatograms <- get_chromatograms(mzml)

# Plot if available
if (length(chromatograms) > 0) {
  chrom <- chromatograms[[1]]
  plot(chrom$time, chrom$intensity,
    type = "l",
    xlab = "Time", ylab = "Intensity",
    main = chrom$chromatogram_type
  )
}
```

## Spectrum Indexing for Memory-Efficient Random Access

For files larger than 10MB, `mzmlr` automatically builds a spectrum index that enables memory-efficient random access without loading the entire XML document into memory. The index stores byte positions of each spectrum, allowing direct seeking to specific spectra:

```{r indexing, eval = FALSE}
# Index is built automatically for files > 10MB
mzml <- read_mzml("large_file.mzML")

# Check if index was built
if (!is.null(mzml$spectrum_index)) {
  cat("Index with", mzml$spectrum_index$n_spectra, "spectra\n")
}

# Force index building for smaller files
mzml <- read_mzml("small_file.mzML", build_index = TRUE)

# Disable index building if not needed
mzml <- read_mzml("file.mzML", build_index = FALSE)
```

When an index is available, `get_spectrum()` reads only the requested spectrum directly from disk:

```{r indexed-reading, eval = FALSE}
# Efficient random access - only reads the requested spectrum
spec1 <- get_spectrum(mzml, 1)
spec1000 <- get_spectrum(mzml, 1000)

# Batch reading also benefits from the index
specs <- get_spectra(mzml, indices = 1:100)
```

## Precision Control

The package provides options to control numeric precision for memory efficiency:

```{r precision, eval = FALSE}
# Default: convert intensities to float32 (~7 significant digits)
# This reduces memory usage by 50% compared to double precision
spec <- get_spectrum(mzml, 1)  # intensity_precision = 0 by default

# Keep full double precision for intensities
spec_full <- get_spectrum(mzml, 1, intensity_precision = NULL)

# Round to specific number of decimal places
spec_rounded <- get_spectrum(mzml, 1, intensity_precision = 2)

# Same options apply to retention times
spec <- get_spectrum(mzml, 1, time_precision = 2)  # default
spec_full_time <- get_spectrum(mzml, 1, time_precision = NULL)
```

## Memory-Efficient Processing

For large files, process spectra in batches to avoid memory issues:

```{r memory-efficient, eval = FALSE}
# Get total count without loading data
info <- get_file_info(mzml)
n_spectra <- info$spectrum_count

# Process in batches using the index for efficient access
batch_size <- 100
for (start in seq(1, n_spectra, by = batch_size)) {
  end <- min(start + batch_size - 1, n_spectra)
  batch <- get_spectra(mzml, indices = start:end)

  # Process batch
  # ...

  # Clear memory
  rm(batch)
  gc()
}
```

## Working with Binary Data

The package automatically handles:
- Base64 decoding
- zlib decompression
- 32-bit float and 64-bit double conversion

```{r binary-data, eval = FALSE}
# All binary data handling is automatic
spec <- get_spectrum(mzml, 1)

# Data is ready for analysis
summary(spec$mz)
summary(spec$intensity)
```

## Error Handling

The package provides informative error messages using the `cli` package:

```{r error-handling, eval = FALSE}
tryCatch(
  {
    mzml <- read_mzml("nonexistent.mzML")
  },
  error = function(e) {
    cat("Error:", conditionMessage(e), "\n")
  }
)
```

Common errors include:
- **File not found**: When the specified path doesn't exist
- **Validation failed**: When the mzML file doesn't conform to the schema
- **Index out of range**: When requesting a spectrum/chromatogram that doesn't exist
- **XML parsing error**: When the file is not well-formed XML

## Troubleshooting

### Tests hanging on Windows

If tests appear to hang indefinitely on Windows, this may be related to file I/O issues. The current version uses binary mode file reading to avoid line-ending issues between Windows (`\r\n`) and Unix (`\n`) systems.

### Memory issues with large files

1. Enable spectrum indexing for files >10MB (automatic by default)
2. Use `intensity_precision = 0` for float32 conversion
3. Process spectra in batches instead of loading all at once
4. Call `gc()` after processing large batches

### Slow validation

Disable validation if you trust your input files:
```r
mzml <- read_mzml("file.mzML", validate = FALSE)
```

## Advanced Features

### Custom Schema Validation

```{r custom-schema, eval = FALSE}
# Use a custom XSD schema for validation
result <- validate_mzml("file.mzML",
  schema_path = "path/to/custom.xsd"
)
```

### Disabling Validation

For faster loading when validation is not needed:

```{r no-validation, eval = FALSE}
mzml <- read_mzml("file.mzML", validate = FALSE)
```

## See Also

- [HUPO-PSI mzML specification](https://www.psidev.info/mzml)
- [mzML GitHub repository](https://github.com/HUPO-PSI/mzML)
- [mzR package](https://rdrr.io/cran/mzR/)
